To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED

### 2023-09-27 manuscript/manuscript.tex
starting to write:

---
title: "To Batch or Not to Batch: Sequential vs. Batched Testing Strategies in the ED"
authors:
- familyname: Jameson
  othernames: Jacob
  affiliation: Interfaculty Initiative in Health Policy, Harvard University
  email: jacobjameson@g.harvard.edu
- familyname: Saghafian
  othernames:  Soroush
  affiliation: Kennedy School of Government, Harvard University 
  email: soroush_saghafian@hks.harvard.edu
- familyname: Huckman
  othernames:  Robert
  affiliation: Harvard Business School 
  email: rhuckman@hbs.edu
- familyname: Hodgson
  othernames:  Nicole
  affiliation: Mayo Clinic
  email: Hodgson.Nicole@mayo.edu
keywords:
  - Emergency Department, Operational Effeciency, Diagnostic Testing
abstract: |
  This paper focuses on analyzing sequential versus batched testing strategies in an emergency department (ED) at Mayo Clinic Arizona, with respect to their associations impacts on patient length of stay, hospital readmission, and healthcare resource utilization. A theoretical model was developed and tested to identify patient and hospital features that may influence the decision to batch or sequentially order tests, such as patient complexity, physician experience, occupancy level and complexity. Finally, we performed retrospective analysis of ED operational data to investigate the impact of batching on our key outcomes. The overall result was that batch ordering tests was associated with greater patient length of stay and resource utilization, even when we control for variation in patient complexity, physician experience, and hospital occupancy.
acknowledgment: |
  The authors would like to thank ..
bibliography: references.bib
blind: false
output: rticles::informs_article
header-includes:
  - \usepackage{booktabs}
  - \usepackage{siunitx}
  - \usepackage{tablefootnote}
  - \usepackage{longtable}
  - \usepackage{threeparttable}
  - \usepackage{natbib}
---

# Introduction {#sec:1}

Healthcare delivery, particularly in the emergency department (ED), is a delicate balance that involves ensuring optimal patient outcomes while optimizing resource utilization. Achieving these twin goals requires timely and accurate diagnosis, which in turn enables prompt and appropriate treatment, consequently improving patient prognosis and reducing the likelihood of adverse events. Furthermore, efficient patient discharge from the ED can help alleviate overcrowding, a severe issue with potential consequences including higher complication rates and increased mortality @bernstein2009.

One important factor that can impact the speed and effectiveness of diagnosis in the ED is the availability and performance of diagnostic tests @naseim2015. A variety of diagnostic tests are used in the ED, including laboratory tests, imaging studies, and specialized tests such as electrocardiograms (ECGs) and point-of-care (POC) testing. These tests can provide valuable information about a patient’s condition and help to guide treatment decisions.

A critical question in this context pertains to whether physicians in the ED should batch order diagnostic tests or order them sequentially. This decision essentially represents a tradeoff between reducing patient length of stay and risk of over-testing. Over-testing, or performing unnecessary tests, can lead to increased costs, unnecessary patient anxiety, and potential harm from follow-up of false-positive results @koch2018. Conversely, keeping a patient for an extended time to perform all possible tests could lead to ED overcrowding, an issue associated with severe consequences, as mentioned earlier. Instead, what is needed is a reasonable balance between the number of diagnostic tests performed and the total time the patient is kept in the ED before either being admitted or discharged. Several studies have demonstrated that optimizing the ED patient flow process can result in significant improvements @saghafian2015, however, research surrounding test ordering strategies to improve the patient flow processes remains limited.

@@

We start by evaluating the reduced-form effects of ED provider batch ordering tendency on downstream patient outcomes and turnaround time. We find that practice variation as captured by physician batch-ordering tendency has large and significant consequences. Being treated by a provider in the top decile of the leniency distribution, compared to being treated by someone in the bottom decile [INSERT RESULTS].

@@

# Data and Definitions {#sec:2}

Our analytical lens is focused on the Emergency Department at the Mayo Clinic of Arizona, a distinguished tertiary care establishment. During our study's timeframe, the ED recorded an annual visitation of approximately 41,000 patients. The department is singularly staffed by board-eligible or board-certified emergency physicians, abstaining from the services of nurse practitioners or physician assistants. A notable observation was that residents in rotation oversaw a low fraction, roughly 10%, of the patient volume. Comprehensive patient data, encompassing demographics, chief complaints, vital signs, emergency severity, length of stay, and resource utilization metrics, were meticulously logged during the study period.

## Sample Construction

@@

- Lab/Image Batch: The first test ordered for a patient was either an imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an additional imaging or lab test order within 5 minutes.

- Image/Image Batch: The first test ordered for a patient was an imaging test (CT scan, X-ray, Ultrasound), followed by an additional imaging test order within 5 minutes.

- Any Batch: This refers to instances where either a Lab/Image Batch or an Image/Image Batch occurred.

@@

## Statistics

Table 1 provides a comprehensive breakdown of the patient encounters involving diagnostic tests ordered either individually (Sequential) or as part of a group (Batch) during an ED encounter from our baseline sample. Around 43% of patient encounters involved diagnostic tests ordered as part of a batch, the majority of which (60%) included two tests.

@@

![Frequency of Sample Cases and Batches](../Manuscript/figures/fig1.png){width=40%, height=30%}

## Institutional Details on Patient-Physician Assignment

@@

# Empirical Strategy {#sec:3}

Our study employs the ``judges design", a quasi-random assignment strategy widely used in literature to investigate causal effects. Typically, this design exploits variations in sentencing leniency among judges working in the same court. We adopt a similar approach by examining batching variations across physicians working in the same emergency department. This method enables us to identify the causal impact of being assigned to different types of physicians with differing batch-ordering tendencies.

To measure physician batch tendency, we use the physician's residualized leave-out average batch rate. This measure is derived from two steps following the approaches taken by @doyle2015measuring, @dobbie2018effects, and @eichmeyer2022pathways. First, we obtain residuals from a regression model, which includes all ED encounters in our sample period.

@@

Where $Batched_{i,t}$ is a dummy variable equal to one if patient $i$ had their diagnostic tests batched on encounter that took place on data $t$. Fixed effects include year-month fixed effects, $\alpha_{ym}$, to control for time and seasonal variation in batching, such as hospital-specific policies (e.g. initiatives to eliminate excess testing) or  seasonality in ED visits. We also control for ``shift-level" variations that include both physician scheduling and patient arrival with day of week-time of day fixed effects, $\alpha_{dt}$. As stated earlier, these controls are what is required for our quasi-random assignment assumption. Under the assumption that we have captured the observables under which quasi-random assignment occurs in the ED, the unexplained variation-- the physician’s contribution-- resides in the error term, $\varepsilon_{i,t}$.

In step two, the leniency measure for patient $i$ seen by physician $j$ is computed as the average residual across all other patients seen by the physician that year:

@@

where $\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}$ is the residual from equation (1); $J$ is the set of all ED encounters treated by physician $j$; and $N_{-i,j} = |\{J \backslash i\}|$, the number of cases that physician has seen that year, excluding patient $i$. This leave-out mean eliminates the mechanical bias that stems from patient $i$'s own case entering into the instrument. The measure is interpreted as the average (leave-out) batch rate of patient $i$’s physician, relative to other physicians in that hospital-year-month, hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic variation in their propensity to batch order diagnostic tests. Figure 2 graphs the histogram of the leniency measure along the x-axis and the left y-axis. A local-linear regression of the fitted probability of batched ordering on tendency after residualizing is overlaid and displayed on the right y-axis. Table 2 presents the same ``first stage" in a regression table: being assigned to a 10 pp more lenient physician is associated with a 17 pp increase in the likelihood of being prescribed an opioid in the ED. The F-statistic is 25 when all controls and fixed effects are included. The coefficient is greater than one because all emergency visits are used to construct the leniency instrument, while the first stage is calculated using the baseline sample only, which excludes the rarely prescribed diagnoses.

@@

This reduced form will allow us to check that our instrument is a strong instrument. To study the effects of test batching in the ED on an outcome $Y_i$, we estimate the following 2SLS equations using our baseline sample:

@@

Where $Y_i$ represents our main outcomes of interest: length of stay, 72 hour readmission, and resource utilization.

@@

and $X_i$ is the same as in the reduced-form approach. $Batched_i$ variable suffers from potential endogeneity concerns. For example, injury severity may be unobserved and correlated with need to run multiple tests, which in turn also affects length of stay. Hence, we instrument $Batched_i$ with the assigned physician $j$'s underlying tendency to batch, $Tendency_{i,j}^{phys}$ We cluster robust standard errors at the physician level to account for the assignment process of patients to physicians.

### Identifying Assumptions

The reduced-form approach delivers an unbiased estimate of the causal effect of being treated by a higher tendency to batch physician, if assignment of patients to ED physicians is random, conditional on seasonality, shift, and diagnosis (``conditional independence"). One can test for whether batch-likely physicians are routinely assigned patients with particular characteristics. Figure 4 shows evidence of balance on observables. The left panel is the output of a regression of Prescribed on observables including patient demographics, priority groups, and ED diagnoses. Unsurprisingly, many observables predict length of stay and 72 hour return. The right panel displays the coefficients from a regression of physician tendency instrument on the same set of observables. With so many coefficients, we fail to reject that the instrument violates conditional independence (the joint F-stat drops by two orders of magnitude, from 280 to 2.4). In fact, some of the variables that are significant in the second column go in the opposite direction of the decision to batch

The residualization in equation (1) controls for more controls than required to achieve quasi-random assignment; they are included for statistical precision in measuring physician tendency to batch. In Appendix Figure [XXX], residualizing for only seasonality and shift (left panel) and also diagnosis (right panel) in equation (1). This lends support to our assumption that assignment in the ED is nearly random at the seasonality, shift  level.

@@

## Results

### Reduced-Form Results

@@

# Conclusion

@@

::: {.APPENDIX latex="{General appendix}"}

@@

\begin{table}[ht]
\centering
\caption{First Stage over Subsamples} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & 0.761*** & 0.00 \\ 
  2 & Upper Respiratory Symptoms & 1.406*** & 0.00 \\ 
  3 & Abnormal Test Results & 1.036*** & 0.00 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.831*** & 0.00 \\ 
  5 & Shortness of Breath & 1.258*** & 0.00 \\ 
  6 & Gastrointestinal Issues & 1.043*** & 0.00 \\ 
  7 & Chest Pain & 0.866*** & 0.00 \\ 
  8 & Dizziness/Lightheadedness/Syncope & 1.668*** & 0.00 \\ 
  9 & Neurological Issue & 0.723*** & 0.00 \\ 
  10 & Fatigue and Weakness & 1.651*** & 0.00 \\ 
  11 & Back or Flank Pain & 0.752*** & 0.01 \\ 
  12 & Abdominal Complaints & 1.109*** & 0.00 \\ 
  13 & Fevers, Sweats or Chills & 1.23*** & 0.00 \\ 
  14 & Skin Complaints & 0.448* & 0.07 \\ 
   \hline
\end{tabular}
\end{table}
:::

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Author template for INFORMS Journal on Data Science (ijds) [interim solution; new styles under construction]
%% Mirko Janc, Ph.D., INFORMS, mirko.janc@informs.org
%% ver. 0.91, March 2015 - updated November 2020 by Matthew Walls, matthew.walls@informs.org
%% Adapted for rticles by Rob J Hyndman Rob.Hyndman@monash.edu. Dec 2021
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[,,nonblindrev]{informs}

\OneAndAHalfSpacedXI
%%\OneAndAHalfSpacedXII % Current default line spacing
%%\DoubleSpacedXII
%%\DoubleSpacedXI

%% BEGIN MY ADDITIONS %%
\usepackage{hyperref}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{booktabs}
\usepackage{siunitx}
\usepackage{tablefootnote}
\usepackage{longtable}
\usepackage{threeparttable}
\usepackage{natbib}

%% END MY ADDITIONS %%

% Natbib setup for author-year style
\usepackage{natbib}
 \bibpunct[, ]{(}{)}{,}{a}{}{,}%
 \def\bibfont{\small}%
 \def\bibsep{\smallskipamount}%
 \def\bibhang{24pt}%
 \def\newblock{\ }%
 \def\BIBand{and}%

%% Setup of theorem styles. Outcomment only one.
%% Preferred default is the first option.
\TheoremsNumberedThrough     % Preferred (Theorem 1, Lemma 1, Theorem 2)
%\TheoremsNumberedByChapter  % (Theorem 1.1, Lema 1.1, Theorem 1.2)
\ECRepeatTheorems

%% Setup of the equation numbering system. Outcomment only one.
%% Preferred default is the first option.
\EquationsNumberedThrough    % Default: (1), (2), ...
%\EquationsNumberedBySection % (1.1), (1.2), ...

% For new submissions, leave this number blank.
% For revisions, input the manuscript number assigned by the on-line
% system along with a suffix ".Rx" where x is the revision number.
\MANUSCRIPTNO{}

%%%%%%%%%%%%%%%%
\begin{document}
%%%%%%%%%%%%%%%%

% Outcomment only when entries are known. Otherwise leave as is and
%   default values will be used.
%\setcounter{page}{1}
%\VOLUME{00}%
%\NO{0}%
%\MONTH{Xxxxx}% (month or a similar seasonal id)
%\YEAR{0000}% e.g., 2005
%\FIRSTPAGE{000}%
%\LASTPAGE{000}%
%\SHORTYEAR{00}% shortened year (two-digit)
%\ISSUE{0000} %
%\LONGFIRSTPAGE{0001} %
%\DOI{10.1287/xxxx.0000.0000}%

% Author's names for the running heads
% Sample depending on the number of authors;
\RUNAUTHOR{%
Jameson, Saghafian, Huckman
 and Hodgson
}
% \RUNAUTHOR{Jones and Wilson}
% \RUNAUTHOR{Jones, Miller, and Wilson}
% \RUNAUTHOR{Jones et al.} % for four or more authors
% Enter authors following the given pattern:
%\RUNAUTHOR{}

\RUNTITLE{To Batch or Not to Batch: Sequential vs.~Batched Testing
Strategies in the ED}

\TITLE{To Batch or Not to Batch: Sequential vs.~Batched Testing
Strategies in the ED}

\ARTICLEAUTHORS{%
\AUTHOR{Jacob Jameson}
\AFF{Interfaculty Initiative in Health Policy, Harvard
University, \EMAIL{\href{mailto:jacobjameson@g.harvard.edu}{\nolinkurl{jacobjameson@g.harvard.edu}}}}

\AUTHOR{Soroush Saghafian}
\AFF{Kennedy School of Government, Harvard
University, \EMAIL{\href{mailto:soroush_saghafian@hks.harvard.edu}{\nolinkurl{soroush\_saghafian@hks.harvard.edu}}}}

\AUTHOR{Robert Huckman}
\AFF{Harvard Business
School, \EMAIL{\href{mailto:rhuckman@hbs.edu}{\nolinkurl{rhuckman@hbs.edu}}}}

\AUTHOR{Nicole Hodgson}
\AFF{Mayo
Clinic, \EMAIL{\href{mailto:Hodgson.Nicole@mayo.edu}{\nolinkurl{Hodgson.Nicole@mayo.edu}}}}

%
}

\ABSTRACT{This paper focuses on analyzing sequential versus batched
testing strategies in an emergency department (ED) at Mayo Clinic
Arizona, with respect to their associations impacts on patient length of
stay, hospital readmission, and healthcare resource utilization. A
theoretical model was developed and tested to identify patient and
hospital features that may influence the decision to batch or
sequentially order tests, such as patient complexity, physician
experience, occupancy level and complexity. Finally, we performed
retrospective analysis of ED operational data to investigate the impact
of batching on our key outcomes. The overall result was that batch
ordering tests was associated with greater patient length of stay and
resource utilization, even when we control for variation in patient
complexity, physician experience, and hospital occupancy.}

\KEYWORDS{Emergency Department, Operational Effeciency, Diagnostic
Testing}

\maketitle

\hypertarget{sec:1}{%
\section{Introduction}\label{sec:1}}

Healthcare delivery, particularly in the emergency department (ED), is a
delicate balance that involves ensuring optimal patient outcomes while
optimizing resource utilization. Achieving these twin goals requires
timely and accurate diagnosis, which in turn enables prompt and
appropriate treatment, consequently improving patient prognosis and
reducing the likelihood of adverse events. Furthermore, efficient
patient discharge from the ED can help alleviate overcrowding, a severe
issue with potential consequences including higher complication rates
and increased mortality \citet{bernstein2009}.

One important factor that can impact the speed and effectiveness of
diagnosis in the ED is the availability and performance of diagnostic
tests \citet{naseim2015}. A variety of diagnostic tests are used in the
ED, including laboratory tests, imaging studies, and specialized tests
such as electrocardiograms (ECGs) and point-of-care (POC) testing. These
tests can provide valuable information about a patient's condition and
help to guide treatment decisions.

A critical question in this context pertains to whether physicians in
the ED should batch order diagnostic tests or order them sequentially.
This decision essentially represents a tradeoff between reducing patient
length of stay and risk of over-testing. Over-testing, or performing
unnecessary tests, can lead to increased costs, unnecessary patient
anxiety, and potential harm from follow-up of false-positive results
\citet{koch2018}. Conversely, keeping a patient for an extended time to
perform all possible tests could lead to ED overcrowding, an issue
associated with severe consequences, as mentioned earlier. Instead, what
is needed is a reasonable balance between the number of diagnostic tests
performed and the total time the patient is kept in the ED before either
being admitted or discharged. Several studies have demonstrated that
optimizing the ED patient flow process can result in significant
improvements \citet{saghafian2015}, however, research surrounding test
ordering strategies to improve the patient flow processes remains
limited.

@@

We start by evaluating the reduced-form effects of ED provider batch
ordering tendency on downstream patient outcomes and turnaround time. We
find that practice variation as captured by physician batch-ordering
tendency has large and significant consequences. Being treated by a
provider in the top decile of the leniency distribution, compared to
being treated by someone in the bottom decile {[}INSERT RESULTS{]}.

@@

\hypertarget{sec:2}{%
\section{Data and Definitions}\label{sec:2}}

Our analytical lens is focused on the Emergency Department at the Mayo
Clinic of Arizona, a distinguished tertiary care establishment. During
our study's timeframe, the ED recorded an annual visitation of
approximately 41,000 patients. The department is singularly staffed by
board-eligible or board-certified emergency physicians, abstaining from
the services of nurse practitioners or physician assistants. A notable
observation was that residents in rotation oversaw a low fraction,
roughly 10\%, of the patient volume. Comprehensive patient data,
encompassing demographics, chief complaints, vital signs, emergency
severity, length of stay, and resource utilization metrics, were
meticulously logged during the study period.

\hypertarget{sample-construction}{%
\subsection{Sample Construction}\label{sample-construction}}

@@

\begin{itemize}
\item
  Lab/Image Batch: The first test ordered for a patient was either an
  imaging or lab test (CT scan, X-ray, Ultrasound, Lab), followed by an
  additional imaging or lab test order within 5 minutes.
\item
  Image/Image Batch: The first test ordered for a patient was an imaging
  test (CT scan, X-ray, Ultrasound), followed by an additional imaging
  test order within 5 minutes.
\item
  Any Batch: This refers to instances where either a Lab/Image Batch or
  an Image/Image Batch occurred.
\end{itemize}

@@

\hypertarget{statistics}{%
\subsection{Statistics}\label{statistics}}

Table 1 provides a comprehensive breakdown of the patient encounters
involving diagnostic tests ordered either individually (Sequential) or
as part of a group (Batch) during an ED encounter from our baseline
sample. Around 43\% of patient encounters involved diagnostic tests
ordered as part of a batch, the majority of which (60\%) included two
tests.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth,height=0.3\textheight]{../Manuscript/figures/fig1.png}
\caption{Frequency of Sample Cases and Batches}
\end{figure}

\hypertarget{institutional-details-on-patient-physician-assignment}{%
\subsection{Institutional Details on Patient-Physician
Assignment}\label{institutional-details-on-patient-physician-assignment}}

@@

\hypertarget{sec:3}{%
\section{Empirical Strategy}\label{sec:3}}

Our study employs the ``judges design'', a quasi-random assignment
strategy widely used in literature to investigate causal effects.
Typically, this design exploits variations in sentencing leniency among
judges working in the same court. We adopt a similar approach by
examining batching variations across physicians working in the same
emergency department. This method enables us to identify the causal
impact of being assigned to different types of physicians with differing
batch-ordering tendencies.

To measure physician batch tendency, we use the physician's residualized
leave-out average batch rate. This measure is derived from two steps
following the approaches taken by \citet{doyle2015measuring},
\citet{dobbie2018effects}, and \citet{eichmeyer2022pathways}. First, we
obtain residuals from a regression model, which includes all ED
encounters in our sample period.

@@

Where \(Batched_{i,t}\) is a dummy variable equal to one if patient
\(i\) had their diagnostic tests batched on encounter that took place on
data \(t\). Fixed effects include year-month fixed effects,
\(\alpha_{ym}\), to control for time and seasonal variation in batching,
such as hospital-specific policies (e.g.~initiatives to eliminate excess
testing) or seasonality in ED visits. We also control for
``shift-level'' variations that include both physician scheduling and
patient arrival with day of week-time of day fixed effects,
\(\alpha_{dt}\). As stated earlier, these controls are what is required
for our quasi-random assignment assumption. Under the assumption that we
have captured the observables under which quasi-random assignment occurs
in the ED, the unexplained variation-- the physician's contribution--
resides in the error term, \(\varepsilon_{i,t}\).

In step two, the leniency measure for patient \(i\) seen by physician
\(j\) is computed as the average residual across all other patients seen
by the physician that year:

@@

where \(\hat{e}_{i',k} = \hat{Batch}_{i',k} - Batch_{i',k}\) is the
residual from equation (1); \(J\) is the set of all ED encounters
treated by physician \(j\); and \(N_{-i,j} = |\{J \backslash i\}|\), the
number of cases that physician has seen that year, excluding patient
\(i\). This leave-out mean eliminates the mechanical bias that stems
from patient \(i\)'s own case entering into the instrument. The measure
is interpreted as the average (leave-out) batch rate of patient \(i\)'s
physician, relative to other physicians in that hospital-year-month,
hospital-day of week-time of day.

We document that the Mayo Clinic ED physicians exhibit wide, systematic
variation in their propensity to batch order diagnostic tests. Figure 2
graphs the histogram of the leniency measure along the x-axis and the
left y-axis. A local-linear regression of the fitted probability of
batched ordering on tendency after residualizing is overlaid and
displayed on the right y-axis. Table 2 presents the same ``first stage''
in a regression table: being assigned to a 10 pp more lenient physician
is associated with a 17 pp increase in the likelihood of being
prescribed an opioid in the ED. The F-statistic is 25 when all controls
and fixed effects are included. The coefficient is greater than one
because all emergency visits are used to construct the leniency
instrument, while the first stage is calculated using the baseline
sample only, which excludes the rarely prescribed diagnoses.

@@

This reduced form will allow us to check that our instrument is a strong
instrument. To study the effects of test batching in the ED on an
outcome \(Y_i\), we estimate the following 2SLS equations using our
baseline sample:

@@

Where \(Y_i\) represents our main outcomes of interest: length of stay,
72 hour readmission, and resource utilization.

@@

and \(X_i\) is the same as in the reduced-form approach. \(Batched_i\)
variable suffers from potential endogeneity concerns. For example,
injury severity may be unobserved and correlated with need to run
multiple tests, which in turn also affects length of stay. Hence, we
instrument \(Batched_i\) with the assigned physician \(j\)'s underlying
tendency to batch, \(Tendency_{i,j}^{phys}\) We cluster robust standard
errors at the physician level to account for the assignment process of
patients to physicians.

\hypertarget{identifying-assumptions}{%
\subsubsection{Identifying Assumptions}\label{identifying-assumptions}}

The reduced-form approach delivers an unbiased estimate of the causal
effect of being treated by a higher tendency to batch physician, if
assignment of patients to ED physicians is random, conditional on
seasonality, shift, and diagnosis (``conditional independence''). One
can test for whether batch-likely physicians are routinely assigned
patients with particular characteristics. Figure 4 shows evidence of
balance on observables. The left panel is the output of a regression of
Prescribed on observables including patient demographics, priority
groups, and ED diagnoses. Unsurprisingly, many observables predict
length of stay and 72 hour return. The right panel displays the
coefficients from a regression of physician tendency instrument on the
same set of observables. With so many coefficients, we fail to reject
that the instrument violates conditional independence (the joint F-stat
drops by two orders of magnitude, from 280 to 2.4). In fact, some of the
variables that are significant in the second column go in the opposite
direction of the decision to batch

The residualization in equation (1) controls for more controls than
required to achieve quasi-random assignment; they are included for
statistical precision in measuring physician tendency to batch. In
Appendix Figure {[}XXX{]}, residualizing for only seasonality and shift
(left panel) and also diagnosis (right panel) in equation (1). This
lends support to our assumption that assignment in the ED is nearly
random at the seasonality, shift level.

@@

\hypertarget{results}{%
\subsection{Results}\label{results}}

\hypertarget{reduced-form-results}{%
\subsubsection{Reduced-Form Results}\label{reduced-form-results}}

@@

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

@@

\begin{APPENDIX}{General appendix}

@@

\begin{table}[ht]
\centering
\caption{First Stage over Subsamples} 
\begin{tabular}{llrr}
  \hline
 & CHIEF\_COMPLAINT & Coefficient\_Starred & P\_value \\ 
  \hline
1 & Extremity Complaints & 0.761*** & 0.00 \\ 
  2 & Upper Respiratory Symptoms & 1.406*** & 0.00 \\ 
  3 & Abnormal Test Results & 1.036*** & 0.00 \\ 
  4 & Falls, Motor Vehicle Crashes, Assaults, and Trauma & 0.831*** & 0.00 \\ 
  5 & Shortness of Breath & 1.258*** & 0.00 \\ 
  6 & Gastrointestinal Issues & 1.043*** & 0.00 \\ 
  7 & Chest Pain & 0.866*** & 0.00 \\ 
  8 & Dizziness/Lightheadedness/Syncope & 1.668*** & 0.00 \\ 
  9 & Neurological Issue & 0.723*** & 0.00 \\ 
  10 & Fatigue and Weakness & 1.651*** & 0.00 \\ 
  11 & Back or Flank Pain & 0.752*** & 0.01 \\ 
  12 & Abdominal Complaints & 1.109*** & 0.00 \\ 
  13 & Fevers, Sweats or Chills & 1.23*** & 0.00 \\ 
  14 & Skin Complaints & 0.448* & 0.07 \\ 
   \hline
\end{tabular}
\end{table}

\end{APPENDIX}

% Appendix here
% Options are (1) APPENDIX (with or without general title) or
%             (2) APPENDICES (if it has more than one unrelated sections)
% Outcomment the appropriate case if necessary
%
% \begin{APPENDIX}{<Title of the Appendix>}
% \end{APPENDIX}
%
%   or
%
% \begin{APPENDICES}
% \section{<Title of Section A>}
% \section{<Title of Section B>}
% etc
% \end{APPENDICES}

% Acknowledgments here
\ACKNOWLEDGMENT{The authors would like to thank ..}

\bibliographystyle{informs2014}
\bibliography{references.bib}

\end{document}
